\documentclass[a4paper]{article}
\usepackage{cmap}
\usepackage[T2A]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb,amsmath,amsfonts}
\usepackage{csquotes}
\usepackage{titling}
\usepackage[useregional]{datetime2}
\usepackage{fullpage}

\pagestyle{empty}

\title{Practice 1: Hartley's formula}
\date{\DTMdate{2024-10-09}}

\preauthor{}
\postauthor{}
\author{}
\setlength{\droptitle}{-50pt}

\begin{document}

\maketitle

\begin{enumerate}
  \item How many yes/no questions need to be asked to determine a number from 1 to \(N\), if the person who thought of the number is allowed to lie once and the questions must be written down in advance?
  Provide an upper estimate.

  \item How many comparisons are needed to sort \(N\) stones by weight (all stones have different weights)?
  Find the exact answer to this question for \(N = 2, 3, 4, 5\).

  \item Given 20 coins, one of which is fake and lighter than the others.
  What is the minimum number of weighings on a balance scale required to determine the fake coin?

  \item Given 12 coins, one of which is fake (with an unknown relative weight). Can it be found in 3 weighings?

  \item Given 15 coins, one of which is fake (with an unknown relative weight). Can it be found in 3 weighings if determining the relative weight of the coin is not required?

  \item Given 14 coins, one of which is fake (with an unknown relative weight). Can it be found in 3 weighings if determining the relative weight of the coin is not required?

  \item How much information is contained in a three-digit number?

  \item Prove that the probability of compressing a binary file (any data) by at least one byte using an archiver does not exceed \(\frac{1}{256}\).

  \item Among \(n\) seemingly identical stones, one is radioactive. A Geiger counter is available to determine whether there is a radioactive stone in any group of stones. Prove that it is necessary and sufficient to use the Geiger counter \(\log n\) times to find the radioactive stone.

\end{enumerate}

\end{document}
